Concurrent, Parallel Garbage Collection in Linear Time

### 2016-06-01 finaletd/mainintroduction.tex
algo updated
intro added

This chapter provides the background information regarding the problem discussed in the thesis. Section
~\ref{intro:gc} describes the fundamental idea behind necessity of garbage collection. Advanced readers can skip the section ~\ref{intro:gc}. Section ~\ref{intro:motv} discusses the advantages of the garbage collection and problem statement of the thesis. Section ~\ref{intro:contr} describes all the contributions of this thesis to solve the problem mentioned in section ~\ref{intro:motv}. Section ~\ref{intro:do} helps the readers to understand how the thesis is organized.
\section{Garbage paragraph}
\label{intro:gc}
All programming languages are designed to perform complex computations. Computation contain three main ingredients : Processing unit, algorithm, and memory. Processing unit aka processor is the unit which performs arithmetic, and logical operations required. Algorithm dictates sequence of operation to performed on the data to get the desired output. Algorithm implemented in a programming language is usually referred as program and collection of programs are called software. Memory saves the input, intermediate results and final output of the computation. In this thesis, we are more concerned with object oriented programming languages and object represent user defined dynamically allocated data. Developers write algorithms in a specific programming languages. Programming language organizes the memory consumed by the algorithm. Modern programming languages uses three different memory to organize it. They are stack, static and heap memory. Static memory contains all the global variables used by the program. Stack memory is used by the programming languages to manage static(fixed size) allocation of data along with scope of the data. Heap memory is used to manage the dynamically allocated objects.

Heap memory helps developer to extend the life of dynamically allocated objects. Some object allocated in the heap might get very long life as long as the life of the program running. So these objects have infinite life if programmer did not delete them after their last use. The unwanted objects in the heap occupies the heap memory and might make application run out of memory and exit. In order to avoid psuedo full heap memory errors, programmers determine the life of the dynamically allocated objects and delete them. The dynamically allocated objects can be accessed only by pointers in the static and stack memory. These pointers are also called as \textbf{roots}.

Garbage collection is process of collecting all unwanted objects in the heap memory. It involves finding if all objects allocated in the heap are in use by the program.
\section{Motivation and Objectives}
\label{intro:motv}
The main advantage of the garbage collections are 
\begin{enumerate}
	\item Dangling Pointers
	\item Double free bugs
	\item Memory Leaks
\end{enumerate}
\paragraph{Dangling Pointers:}
	Dangling pointers are those pointers where programmer deleted the object a pointer points to. So these pointers are very dangerous in nature as one does not know what the pointer is pointing to.
\paragraph{Double free bugs:}
	Double free bugs is problem when an objected allocated in the heap is deleted and programmer deletes the object again. The problem with this if a new object is allocated in the space and the delete is called, the new object is deleted without knowledge. This creates some dangling pointers.
\paragraph{memory leaks:}
	The most important of all is that the object is allocated in the heap and it is not used. The unused object is not deleted and it consumes memory which could be reused for other objects.

>>>>

\begin{comment}
This chapter provides the background information regarding the problem discussed in the thesis. Section
~\ref{intro:gc} describes the fundamental idea behind necessity of garbage collection. Advanced readers can skip the section ~\ref{intro:gc}. Section ~\ref{intro:motv} discusses the advantages of the garbage collection and problem statement of the thesis. Section ~\ref{intro:contr} describes all the contributions of this thesis to solve the problem mentioned in section ~\ref{intro:motv}. Section ~\ref{intro:do} helps the readers to understand how the thesis is organized.
\end{comment}
\section{Garbage Collection}
\label{intro:gc}

Processor, algorithm (software), and memory are the trio of computation. Memory is organized by the programming languages and developers can control the life of data allocated in the memory. In this thesis, we abstract all data allocated in the memory as object. In general, we define object as any block of memory that saves any information in the memory and be able to point to other piece of memory by address. Most of the programming supports dynamic memory allocation. To organize memory better, programming languages divide available memory into three types. 
\begin {comment}
All programming languages are designed to perform complex computations. Computation contain three main ingredients : Processing unit, algorithm, and memory. Processing unit i.e., processor is the unit which performs arithmetic, and logical operations required. Algorithm dictates sequence of operation to performed on the data to get the desired output. Algorithm implemented in a programming language is usually referred as program and collection of programs are called software. Memory saves the input, intermediate results and final output of the computation. 
In this thesis, we are more concerned with object oriented programming languages and object represent user defined dynamically allocated data. Developers write algorithms in a specific programming language. Programming language organizes the memory consumed by the algorithm. Modern programming languages uses three different memories to organize it.
\end{comment} 
They are stack, static and heap memory. Static memory contains all the global variables used by the program. Stack memory is used to manage static(fixed size) allocation of object whose scope is well defined. Heap memory is used to manage the dynamically allocated objects which usually has infinite scope unless until mentioned by developers.

Static memory size requirement is computed at the compilation phase and does not contain objects that are unused at point in time. Stack memory is used to manage the function calls and allocation of objects in the function scope. So the size of the stack memory required for the execution cannot be computed at compilation time, but the size of each call frame can be computed. Uncertainty in the size of stack and heap memory is solved by modeling the remaining memory as two sliders for a box. The size of each section is highly customizable. The idea behind the organization is stack memory is filled from lower address of memory and heap memory is filled from the higher address of the memory. When sliders of both memory meets, the program usually crash if additional memory is requested by the memory.

Stack memory is memory of all objects that are actively in use or will be used in the near future. Heap memory is memory of all objects whose scope is unknown to the program. Developers control the lifetime of the objects in the heap memory and deletes them when it is no longer used. Memory management is the term used to describe heap memory management. Developers write program that creates the objects in the heap memory and deletes them when it is of no use. The developer managed memory is referred as manual memory management.

 There are three main issues that happens when heap is manually managed. They are dangling pointers, double free bugs and memory leaks. All of the three issues are deadly and harmful. Some of them produce wrong output to the programs, deletes memory that is in use, and also inefficiently organizes the heap memory. Automatic memory avoids all the above mentioned problems. Apart from all the above mentioned problems, it is extremely difficult to manually manage memory for certain class of programs called concurrent programs. This thesis focuses on designing algorithms to provide finest quality automatic memory management to programming languages runtime systems. The process of automatically managing memory is termed as \textbf{Garbage Collection (GC)}.

\begin{comment}
Heap memory helps developer to extend the life of dynamically allocated objects. Some object allocated in the heap might get very long life as long as the life of the program running. So these objects have infinite life if programmer did not delete them after their last use. The unwanted objects in the heap occupies the heap memory and might make application run out of memory and exit. In order to avoid psuedo full heap memory errors, programmers determine the life of the dynamically allocated objects and delete them. The dynamically allocated objects can be accessed only by pointers in the static and stack memory. These pointers are also called as \textbf{roots}.

\end{comment}
\subsection{How GC's work}
The term garbage refers to the objects allocated in the heap that are not used anymore in the programs. All objects allocated in the memory can referenced by their address. An object $x$ is said to hold a reference to object $y$ when $x$ saves the address of $y$ in it and can always read contents of $y$.To use any object allocated in the heap memory, an object in the stack or static memory must be able to read content in the object through the reference saved in the object itself or through chain of references in multiple object chain.
The objects in the stack and static memory that are considered for defining the use of the object are referred as \textbf{roots}.
So when any root object read contents of an object in the heap memory, then the object cannot be used by the program and should be classified as garbage objects. Dead is an alternative adjective used to describe the objects lost its life and will not be used anymore. Garbage collection is process of collecting all dead objects in the heap memory.

There are two major theoretical technique to identify garbage objects.They are tracing and reference counting. Tracing involves marking all the object in the heap that are in use by going through all the chain of references from roots. Once the marking is done, all unmarked object in the heap is considered as garbage and deleted by the garbage collection program. Reference counting approach attaches a counter to each object that counts number of object that hold reference to it. When the counter reaches zero, the object identifies that it is garbage and deletes itself.

\section{Motivation and Objectives}
\label{intro:motv}
The main advantage of the garbage collections are 
\begin{enumerate}
	\item Dangling Pointers
	\item Double free bugs
	\item Memory Leaks
\end{enumerate}
\paragraph{Dangling Pointers:}
	Dangling pointers are those pointers where programmer deleted the object a pointer points to. So these pointers are very dangerous in nature as one does not know what the pointer is pointing to.
\paragraph{Double free bugs:}
	Double free bugs is problem when an objected allocated in the heap is deleted and programmer deletes the object again. The problem with this if a new object is allocated in the space and the delete is called, the new object is deleted without knowledge. This creates some dangling pointers.
\paragraph{memory leaks:}
	The most important of all is that the object is allocated in the heap and it is not used. The unused object is not deleted and it consumes memory which could be reused for other objects.
